---
title: "Diagnosis vs. Relapse Expression Analysis"
author: "Timothy Keyes"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---
# Executive Summary 

In this report, we analyze the expression of several markers in clustered cell populations between paired diagnostic and relapse samples from AML patients (as well as healthy controls). 

# Analysis


```{r expression_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi = 300, warning = FALSE, message = FALSE)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(ggridges)

# Parameters
input_path <- here::here("data", "AML_matrix_clustered.rds")

CLASSIFIER_POPULATIONS <- 
  c(
    'HSC',
    'MPP',
    'CMP',
    'GMP',
    'MEP',                          
    'Monocyte', 
    'DC', 
    'Macrophage', 
    'Thrombocyte'
  )

#===============================================================================

#source necessary files
source('~/GitHub/aml-cells/scripts/setup/marker_setup.R')
source('~/GitHub/aml-cells/scripts/setup/patient_setup.R')

```

## Basic setup

Algorithm (in words): 

* Read in previously pre-processed data
  * arcsinh transform
  * compensation
  * filtered on `time` variable to remove weird CD34/CD38 "debarcode-gate"
* filter to only include paired samples (with both `Dx` and `Rx` present as well as healthy controls)

```{r}
#### set up global varibles
marker_setup()
patient_setup()

recode_vars <- 
  tibble(
    from = 1:length(CLASSIFIER_POPULATIONS), 
    to = CLASSIFIER_POPULATIONS
  ) %>% 
  deframe()

#read in paired data (and healthy data) 
aml_data <- 
  input_path %>% 
  read_rds() %>%
  mutate(
    patient = 
      patient %>% 
      as.character() %>% 
      str_to_lower() %>% 
      if_else(. == "pastrp", "pasrtp", .), 
    Mah.cluster = 
      recode(Mah.cluster, !!! recode_vars) %>% 
      factor(levels = CLASSIFIER_POPULATIONS)
  ) %>% 
  mutate_at(
    .vars = vars(phenograph.metacluster, FlowSOM.cluster, FlowSOM.metacluster),
    ~ (.) %>% 
      as.character() %>% 
      fct_reorder(., CD34, .fun = mean, .desc = TRUE)
  ) %>% 
  dplyr::filter(patient %in% c(PAIRED_PATIENTS, HEALTHY_CONTROLS)) %>% 
  rename_at(
    vars(everything()), 
    ~ (.) %>% 
      str_to_lower() %>% 
      str_replace_all(pattern = "[:punct:]", replacement = "_")
  )
```

## Cluster to cluster across patients

```{r}
cluster_column <- "mah_cluster"
marker <- "cd45"

clusters_across_patients <- function(cluster_column, marker) {
  
  map(
    levels(aml_data[[cluster_column]]), 
    function(cluster_type) {
      patient_levels <- 
        aml_data %>% 
        dplyr::filter(!! sym(cluster_column) == cluster_type) %>% 
        group_by(
          metacondition = if_else(condition == "Healthy", "Healthy", "Cancer"),
          patient
        ) %>% 
        summarize(median = median(!! sym(marker))) %>% 
        ungroup() %>% 
        arrange(metacondition, median) %>% 
        pull(patient)
      
      aml_data %>% 
        dplyr::filter(!! sym(cluster_column) %in% cluster_type) %>% 
        dplyr::filter(stimulation == "Basal") %>% 
        mutate(
          metacondition = 
            if_else(condition == "Healthy", "Healthy", "Cancer") %>% 
            as.factor(), 
          patient = factor(patient, levels = patient_levels)
        ) %>% 
        ggplot(aes(x = !! sym(marker), y = patient, fill = condition)) + 
        geom_density_ridges(scale = 1, alpha = 0.6) + 
        coord_cartesian(clip = "off") + 
        theme_ridges() + 
        labs(
          title = str_c(
            str_to_upper(marker), 
            " expression in ", 
            cluster_type, 
            " cells across patients"
          ), 
          x = str_c(marker, " expression"), 
          y = NULL,
          fill = NULL
        )
    }
  )
}
```


### CD45
```{r, fig.height=12}
cd45_mah_cluster_plots <- clusters_across_patients("mah_cluster", "cd45")

cd45_mah_cluster_plots %>% 
  walk(print)
```

-----------------------

### CD34

```{r, fig.height=12}
cd34_mah_cluster_plots <- clusters_across_patients("mah_cluster", "cd34")

cd34_mah_cluster_plots %>% 
  walk(print)
```

-----------------------

### CD38 

```{r, fig.height=12}
cd38_mah_cluster_plots <- clusters_across_patients("mah_cluster", "cd38")

cd38_mah_cluster_plots %>% 
  walk(print)
```

-----------------------

### CD90 

```{r, fig.height=12}
cd90_mah_cluster_plots <- clusters_across_patients("mah_cluster", "cd90")

cd90_mah_cluster_plots[1:2] %>% 
  walk(print)
```

-----------------------

### CD93

```{r, fig.height=12}
cd93_mah_cluster_plots <- clusters_across_patients("mah_cluster", "cd93")

cd93_mah_cluster_plots[1:2] %>% 
  walk(print)
```

-----------------------

### CD99 

```{r, fig.height=12}
cd99_mah_cluster_plots <- clusters_across_patients("mah_cluster", "cd99")

cd99_mah_cluster_plots[1:2] %>% 
  walk(print)
```

-----------------------

### pSTAT5

```{r, fig.height=12}
pstat5_mah_cluster_plots <- clusters_across_patients("mah_cluster", "pstat5")

pstat5_mah_cluster_plots[1:3] %>% 
  walk(print)
```

-----------------------

### pSTAT3 

```{r, fig.height=12}
pstat3_mah_cluster_plots <- clusters_across_patients("mah_cluster", "pstat3")

pstat3_mah_cluster_plots[1:3] %>% 
  walk(print)
```

----------------------


#### Conclusions

* Univariate analyses don't indicate any particular marker as "leaping out" as particularly higher or lower in relapse samples relative to healthy samples (to be expected)

* Often, magnitudes and even directions of up- or down-regulation are not consistent between diagnostic-relapse pairs. 

* Marker distributions within a patient and population seem to stay relatively constant for surface markers, but less so for functional markers (again, perhaps to be expected). So, our model should probably not extract the mean or median of the signaling distributions for fear of losing whatever differences are present.

* Classifier population marker distributions can vary quite a bit from patient to patient. 

* Some markers are not very highly expresssed at all. This is evidence that the panel can still be optimized (FYI: Astraea and I have already worked on this quite a bit for AML panel 2.0) 

* Differences across the board are likely to be very nuanced. Perhaps patient-level features will not capture this: should the model operate on the single-cell level? 
