---
title: "Diagnosis vs. Relapse Abundance Analysis"
author: "Timothy Keyes"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

# Executive Summary 

In this analysis, we look at the abundance of different clusters of myeloid cells (in both healthy and AML bone marrow) in paired diagnostic and relapse samples from the same patients. 

# Analysis

```{r abundance_setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi = 300)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
#library(CATALYST)

# Parameters
input_path <- here::here("data", "AML_matrix_clustered.rds")

CLASSIFIER_POPULATIONS <- 
  c(
    'HSC',
    'MPP',
    'CMP',
    'GMP',
    'MEP',                          
    'Monocyte', 
    'DC', 
    'Macrophage', 
    'Thrombocyte'
  )

#===============================================================================

#source necessary files
source('~/GitHub/aml-cells/scripts/setup/marker_setup.R')
source('~/GitHub/aml-cells/scripts/setup/patient_setup.R')

```

### Basic setup

Algorithm (in words): 

* Read in previously pre-processed data
  * arcsinh transform
  * compensation
  * filtered on `time` variable to remove weird CD34/CD38 "debarcode-gate"
* filter to only include paired samples (with both `Dx` and `Rx` present as well as healthy controls)

```{r}
#### set up global varibles
marker_setup()
patient_setup()

recode_vars <- 
  tibble(
    from = 1:length(CLASSIFIER_POPULATIONS), 
    to = CLASSIFIER_POPULATIONS
  ) %>% 
  deframe()

#read in paired data (and healthy data) 
aml_data <- 
  input_path %>% 
  read_rds() %>%
  mutate(
    patient = 
      patient %>% 
      as.character() %>% 
      str_to_lower() %>% 
      if_else(. == "pastrp", "pasrtp", .), 
    Mah.cluster = 
      recode(Mah.cluster, !!! recode_vars) %>% 
      factor(levels = CLASSIFIER_POPULATIONS)
  ) %>% 
  mutate_at(
    .vars = vars(phenograph.metacluster, FlowSOM.metacluster, kmeans.cluster),
    ~ (.) %>% 
      as.character() %>% 
      fct_reorder(., CD34, .fun = mean, .desc = FALSE)
  ) %>% 
  filter(patient %in% c(PAIRED_PATIENTS, HEALTHY_CONTROLS)) %>% 
  rename_at(
    vars(everything()), 
    ~ (.) %>% 
      str_to_lower() %>% 
      str_replace_all(pattern = "[:punct:]", replacement = "_")
  )
```

### Differential abundance visualization

```{r, echo = FALSE}
### function to make plots
cluster_plot <- function(cluster_column) { 
  
  sample_size <- 
    aml_data %>% 
    group_by(condition) %>% 
    summarize(group_size = n_distinct(patient))
  
  abundances <- 
    aml_data %>% 
    group_by_at(vars(one_of("patient", "condition", cluster_column))) %>%
    summarize(cell_count = n()) %>% 
    mutate(cell_prop = cell_count / sum(cell_count)) %>% 
    left_join(sample_size, by = "condition") %>% 
    group_by_at(vars(one_of("condition", cluster_column))) %>% 
    summarize(
      group_size = mean(group_size), 
      sem = sd(cell_prop) / sqrt(group_size), 
      cell_prop = mean(cell_prop), 
    )

  ### Make plots 
  plot <- 
    abundances %>% 
    ggplot(aes(x = !! sym(cluster_column), y = cell_prop, color = condition)) + 
    geom_line(aes(group = condition)) + 
    geom_crossbar(
      aes(ymin = cell_prop - sem, ymax = cell_prop + sem, fill = condition), 
      alpha = 0.3, 
      color = NA
    ) +
    geom_point(shape = 21, fill = "white", size = 3) + 
    scale_y_continuous(
      breaks = scales::breaks_width(width = 0.1), 
      labels = scales::label_percent(accuracy = 1)
    ) +
    scale_color_discrete(
      breaks = c("Healthy", "Dx", "Rx"), 
      labels = c("Healthy", "Diagnosis", "Relapse")
    ) + 
    scale_fill_discrete(
      breaks = c("Healthy", "Dx", "Rx"), 
      labels = c("Healthy", "Diagnosis", "Relapse")
    ) + 
    theme_light() + 
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), 
      legend.position = c(0.07, 0.9), 
      legend.background = element_blank()
    ) + 
    labs(
      x = NULL, 
      y = "Cells in population (%)", 
      fill = NULL, 
      color = NULL, 
      subtitle = 
        "Dots indicate mean across all observations; Crossbars indicate 1 SEM"
    )
  
  plot
}
```

#### Classifier populations

**In aggregate:**

Algorithm (in words): 

* For each patient and diagnostic condition (i.e. "Healthy", "Diagnosis", and "Relapse"), find the abundance of each classifier population (calculated using these markers): 

```{r, echo = FALSE, warning = FALSE, message = FALSE}
tibble(marker = CLASSIFIER.MARKERS) %>% 
  knitr::kable()
```

* Find the mean and standard error of the mean for each population in each condition (across all patients) 
* Plot

```{r, echo = FALSE}
cluster_plot("mah_cluster") 
```

From this plot, as discussed before, we can see that primitive populations (including HSC, MPP, CMP GMP, and MEP) are expanded in the cancer samples relative to healthy patients. However, there don't seem to be many populations that are very different between diagnostic and relapse samples at least in aggregate. However, we can take a closer look by doing a patient-specific analysis...

**Patient by patient:**

```{r, echo = FALSE}
##write function to make plots 
patient_plot <- function(cluster_column) {
  
  abundances <- 
    aml_data %>% 
    group_by_at(vars(one_of("patient", "condition", cluster_column))) %>%
    summarize(cell_count = n()) %>% 
    mutate(cell_prop = cell_count / sum(cell_count))
  
  plot <- 
    abundances %>% 
    ggplot(aes(x = !! sym(cluster_column), y = cell_prop, color = condition)) + 
    geom_line(aes(group = str_c(patient, condition))) +
    geom_point(size = 2) + 
    scale_y_continuous(
      breaks = scales::breaks_width(width = 0.1), 
      labels = scales::label_percent(accuracy = 1)
    ) +
    scale_color_discrete(
      breaks = c("Healthy", "Dx", "Rx"), 
      labels = c("Healthy", "Diagnosis", "Relapse")
    ) + 
    scale_fill_discrete(
      breaks = c("Healthy", "Dx", "Rx"), 
      labels = c("Healthy", "Diagnosis", "Relapse")
    ) + 
    facet_wrap(facets = vars(patient), ncol = 4) + 
    theme_light() + 
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), 
      legend.position = "right", 
      legend.background = element_blank()
    ) + 
    labs(
      x = NULL, 
      y = "Cells in population (%)", 
      fill = NULL, 
      color = NULL, 
      subtitle = 
        "Dots indicate overall abundance in each patient and diagnostic condition."
    )
  
  plot
}
```

```{r, fig.width = 15, fig.height = 10}
patient_plot("mah_cluster")
```

From these plots, a qualitative analysis might suggest that there are 3 main "types" of relationship between `diagnosis` and `relapse` timepoints: 

  1) **Patients who don't seem to change much:** PARBIU, PATGIY, PATJSP, PARWXU, PATHIW
  1) **Patients who shift from 1 main population to another:** PARZUU, PASXVC, PASBEE
  1) **Patients with minor shifts (across multiple populations):** PASGWH, PASVZC, PASRTP, PATLHB
  




#### Phenograph populations

**In aggregate:**

Algorithm (in words): 

* For each patient and diagnostic condition (i.e. "Healthy", "Diagnosis", and "Relapse"), find the abundance of each phenograph population (calculated using all surface markers, as done in the original Phenograph paper).

* Find the mean and standard error of the mean for each population in each condition (across all patients).

* Order the populations by mean `CD34` expression (used as a marker of immaturity in the original Phenograph paper)

* Plot

```{r}
cluster_plot("phenograph_metacluster") + 
    labs(x = "Decreasing CD34 expression →")
```

From this plot, we can see that most phenograph clusters don't have SEM error bars, indicating that they are only present in 1 patient in a given diagnostic condition. For that reason, PhenoGraph clustering may need to be run with alternative parameters if patient-to-patient comparisons are to be made. 

In addition, we can see that there isn't quite as much specificity for expansion in the "immature" compartment (i.e. the CD34-high compartment) compared to the classifier's immature populations. This is likely because CD34 expression is really a "poor man's" way to indicate immaturity, and a more rigorous way of expressing this idea would be to explicitly compute a pseudo-time axis of some kind. 


```{r, fig.width = 20, fig.height = 10, eval = FALSE, echo = FALSE}
# **Patient-by-patient:**

patient_plot("phenograph_metacluster") + 
  labs(x = "Decreasing CD34 expression →")

```

#### FlowSOM populations

**In aggregate:**

Algorithm (in words): 

* For each patient and diagnostic condition (i.e. "Healthy", "Diagnosis", and "Relapse"), find the abundance of each FlowSOM population (calculated using all surface markers).

* Find the mean and standard error of the mean for each population in each condition (across all patients).

* Order the populations by mean `CD34` expression (used as a marker of immaturity in the original Phenograph paper)

* Plot

```{r}
cluster_plot("flowsom_metacluster") + 
  theme(legend.position = c(0.92, 0.9)) + 
  labs(x = "Decreasing CD34 expression →")
```

Here, we can see similar results to that found in the classifier: several populations are expanded in the cancer samples relative to the healthy samples, with 2 of them (8 and 1) reduced in the cancer samples relative to the healthy samples overall. 

Note that a simliar issue w.r.t. using `CD34` expression as a maturity marker is seen as above. 

**Patient-by-patient:**

When we look at similar plots on a patient-by-patient basis, we can see that the same kinds of patterns between `diagnosis` and `relapse` time points within a single patients are observed using FlowSOM clustering as were observed using the classifier populations. 


```{r, fig.width = 15, fig.height = 10}
patient_plot("flowsom_metacluster") + 
  labs(x = "Decreasing CD34 expression →")

```

Once again, our qualitative analysis might suggest that there are 3 main "types" of relationship between `diagnosis` and `relapse` timepoints: 

  2) **Patients who don't seem to change much:** PARBIU, PATGIY, PATJSP, PARWXU, PATHIW
  2) **Patients who shift from 1 main population to another:** PARZUU, PASXVC, PASBEE
  2) **Patients with minor shifts (across multiple populations):** PASGWH, PASVZC, PASRTP, PATLHB
  

### Conclusions

* Abundances of different cell populations are likely to be informative features in a predictive model of relapse, as difference between patients (and between time points) suggest a role in relapse biology 

*  There is variability between patients (i.e. 3 types of patient) re: shifts in population abundance from diagnosis to relapse. This may indicate that a model that can take into account the relationships between populations may capture more relevant information than one that only considers them independently (the current gold standard).

* Pseudo-time calculations among healthy and cancer cell populations may yield a relevant feature for modeling (interaction term, see point above) or may simply help with visualization. 

* Statistics to evaluate the descriptive differences above would be nice. Currently, I am evaluating between the two main approaches that people use for single-cell data: 
  * SAM-like permutation tests.
  * Linear mixed models (i.e. `diffcyt` and `limma` packages). 
