---
title: "Single-cell feature EDA"
author: tkeyes
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi = 300)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)

libraries <- 
  c(
    "tidyverse", 
    "tidymodels", 
    "rlang", 
    "ggthemes", 
    "doParallel", 
    "ggridges"
  )

source(here::here("scripts", "setup", "aml_utils.R")) 
call_libraries(libraries)

# Parameters
set_global_variables(locale = "galaxia")
md_path <- here::here("data-raw", "AML_metadata.xlsx")
tidyTOF_directory <- file.path("~", "GitHub", "tidyTOF")

single_cell_path <- here::here("data", "aml_data_clustered.rds")
patient_features_path <- here::here("data", "mahalanobis_dev_features_10.rds")
feature_path <- here::here("data", "modeling_logistic", "mah_dev_feature_names.rds")
md_path <- here::here("data", "md.rds")


# Sourcing tidyTOF functions
source_tidyTOF(tidyTOF_directory)

# set up aml marker and patient information
marker_setup()
patient_setup()

```

# Setup

## Read in data

```{r}
metadata <- 
  md_path %>% 
  read_rds() %>% 
  transmute(
    patient, 
    age = age_at_diagnosis_in_days, 
    wbc = wbc_at_diagnosis, 
    outcome = first_event %>% str_replace("Censored", "Non-relapse"), 
    fab_category, 
    cns_disease, 
    primary_cytogenetic_code, 
    cytogenetic_complexity, 
    aml0531_risk = 
      case_when(
        inv_16 == "Yes" | t_8_21 == "Yes" ~ "Low", 
        any(
          monosomy_5 == "Yes", 
          monosomy_7 == "Yes", 
          del5q == "Yes", 
          !is.na(flt3_itd_allelic_ratio)
        )                                 ~ "High", 
        TRUE                              ~ "Intermediate"
      )
  )

aml_data <- 
  single_cell_path %>% 
  read_rds() %>% 
  group_by(file_names) %>% 
  slice_sample(prop = 0.5) %>% 
  ungroup()


patient_features <- 
  patient_features_path %>% 
  read_rds() %>% 
  left_join(metadata, by = "patient") %>% 
  filter(!is.na(outcome)) %>%
  replace(list = is.na(.), values = 0)


```


```{r}
metadata %>% 
  count(primary_cytogenetic_code) %>% 
  arrange(desc(n))

metadata %>% 
  count(fab_category) %>% 
  arrange(desc(n))

metadata %>% 
  count(cns_disease) %>% 
  arrange(desc(n))

metadata %>% 
  count(cytogenetic_complexity) %>% 
  arrange(cytogenetic_complexity)

```


## Read in feature names 

```{r}
feature_names <- 
  feature_path %>% 
  read_rds() %>% 
  tibble(original_names = .) %>% 
  mutate(
    cell_population = 
      if_else(
        str_detect(string = original_names, pattern = "_"),
        true = 
          str_extract(
            original_names,
            pattern = str_c(CLASSIFIER_POPULATIONS, collapse = "|")
          ), 
        false = NA_character_
      ), 
    marker = 
      if_else(
        str_detect(string = original_names, pattern = "_"), 
        true = 
          str_extract(
            original_names, 
            pattern = str_c(ALL_MARKERS, collapse = "|")
          ), 
        false = NA_character_
      ), 
    stimulation = 
      if_else(
        str_detect(string = original_names, pattern = "_"), 
        true = 
          str_extract(
            original_names, 
            pattern = str_c(unique(aml_data$stimulation), collapse = "|")
          ), 
        false = NA_character_
      ), 
    feature_type = 
      case_when(
        is.na(cell_population) ~ "patient-level",
        is.na(marker)          ~ "abundance",
        is.na(stimulation)     ~ "surface", 
        TRUE                   ~ "signaling"
      )
  )

feature_names

```

# construct final model

Next, we fit the final model (and only using the features we're interested in). 

```{r}
model_df <- 
  patient_features %>% 
  select(all_of(feature_names$original_names))

predicted_outcome <- 
  logistic_reg(penalty = 0, mixture = 0) %>% 
  set_engine("glmnet") %>% 
  fit_xy(
    x = model_df, 
    y = as.factor(patient_features$outcome)
  ) %>% 
  predict(model_df) %>% 
  pull(.pred_class)
```


```{r}
prediction_summary <- 
  patient_features %>% 
  mutate(
    predicted_outcome = if_else(predicted_outcome == "Non-relapse", "Low", "High"), 
    hybrid_prediction = any(predicted_outcome == "High", aml0531_risk == "High")
  ) %>% 
  group_by(outcome) %>% 
  summarize(
    predicted_correct = 
      sum(
        predicted_outcome == "High" & outcome == "Relapse" | 
          predicted_outcome == "Low" & outcome == "Non-relapse"
      ) / n(), 
    aml0531_correct = 
      sum(
        aml0531_risk == "High" & outcome == "Relapse" | 
          aml0531_risk == "Low" & outcome == "Non-relapse"
      ) / n(), 
    hybrid_correct = 
      sum(
        hybrid_prediction & outcome == "Relapse" | 
          !hybrid_prediction & outcome == "Non-relapse"
      ) / n()
  )


prediction_summary
```


 


# Explore features in diagnostic samples

## Patient-level features

```{r}
feature_names %>% 
  filter(feature_type == "patient-level") %>% 
  pull(original_names) %>% 
  select(.data = patient_features, all_of(.), patient, condition, outcome) %>% 
  filter(condition == "dx") %>% 
  ggplot(aes(x = outcome, y = wbc, fill = outcome)) + 
  geom_violin(draw_quantiles = 0.5) + 
  geom_point(alpha = 0.8)

```

## Abundance features 

```{r}
# feature_names %>% 
#   filter(feature_type == "abundance") %>% 
#   pull(original_names) %>% 
#   select(.data = patient_features, all_of(.), patient, condition, outcome) %>%
  
patient_features %>% 
  filter(condition == "dx") %>% 
  mutate(outcome = fct_rev(outcome)) %>% 
  ggplot(aes(x = outcome, y = Thrombocyte_abundance, fill = outcome)) + 
  geom_violin(draw_quantiles = 0.5) + 
  geom_point(alpha = 0.8) + 
  ggrepel::geom_text_repel(aes(label = fab_category), nudge_x = 0.1) + 
  scale_y_continuous(labels = scales::label_percent())
```

## Surface markers 

```{r}
aml_data <- 
  aml_data %>% 
  left_join(metadata, by = c("patient"))

```


```{r, warning = FALSE, message = FALSE, out.width = "100%"}

plot_surface <- function(cell_population, marker){
  
  patient_order <- 
    aml_data %>% 
    filter(condition == "dx", mahalanobis_cluster == cell_population) %>% 
    group_by(mahalanobis_cluster, patient) %>% 
    filter(n() > 50) %>% 
    group_by(patient, fab_category, outcome) %>% 
    summarize(marker := mean(!!sym(marker))) %>% 
    ungroup() %>% 
    arrange(outcome, desc(fab_category), desc(marker)) %>% 
    pull(patient)
  
  aml_data %>% 
    filter(condition == "dx", mahalanobis_cluster == cell_population) %>% 
    group_by(mahalanobis_cluster, patient) %>% 
    filter(n() > 50) %>% 
    ungroup() %>% 
    mutate(
      patient = factor(patient, levels = patient_order), 
      outcome = fct_rev(outcome)
    ) %>% 
    ggplot(aes_string(x = marker, y = "patient", fill = "outcome")) + 
    geom_density_ridges() + 
    geom_text(
      aes(label = fab_category),
      data = 
        aml_data %>% 
        filter(condition == "dx", mahalanobis_cluster == cell_population) %>% 
        group_by(mahalanobis_cluster, patient) %>% 
        filter(n() > 50) %>% 
        ungroup() %>% 
        mutate(!!marker := max(!!sym(marker))) %>% 
        group_by(patient) %>% 
        slice_head(n = 1), 
      nudge_y = 0.5, 
      size = 3
    ) + 
    labs(
      x = str_c(marker, " expression (arcsinh-counts)"), 
      y = "Density of cells", 
      subtitle = str_c(marker, " in ", cell_population, "-like cells at diagnosis")
    )
}

surface_plots <- 
  feature_names %>% 
  filter(feature_type == "surface") %>%
  transmute(
    plots = map2(.x = cell_population, .y = marker, .f = plot_surface)
  )
  
surface_plots %>% 
  pull(plots) %>% 
  walk(print)
  

```

## Signaling features

```{r, warning = FALSE, message = FALSE, out.width = "100%"}
plot_signaling <- function(cell_population, marker, stimulation){
  
  patient_order <- 
    aml_data %>% 
    filter(condition == "dx", mahalanobis_cluster == cell_population) %>% 
    group_by(mahalanobis_cluster, patient) %>% 
    filter(n() > 50) %>% 
    group_by(patient, fab_category, outcome) %>% 
    summarize(marker := mean(!!sym(marker))) %>% 
    ungroup() %>% 
    arrange(outcome, desc(fab_category), desc(marker)) %>% 
    pull(patient)
  
  aml_data %>% 
    filter(condition == "dx", mahalanobis_cluster == cell_population, stimulation == stimulation) %>% 
    group_by(mahalanobis_cluster, patient) %>% 
    filter(n() > 50) %>% 
    ungroup() %>% 
    mutate(
      patient = factor(patient, levels = patient_order), 
      outcome = fct_rev(outcome)
    ) %>% 
    ggplot(aes_string(x = marker, y = "patient", fill = "outcome")) + 
    geom_density_ridges() + 
    labs(
      x = str_c(marker, " expression (arcsinh-counts)"), 
      y = "Density of cells", 
      subtitle = str_c(marker, " in ", cell_population, "-like cells at diagnosis (Stimulation: ", stimulation, ")")
    )
}

signaling_plots <- 
  feature_names %>% 
  filter(feature_type == "signaling") %>%
  transmute(
    plots = pmap(.l = list(cell_population, marker, stimulation), .f = plot_signaling)
  )
  
signaling_plots %>% 
  pull(plots) %>% 
  walk(print)
```

## Some more depth with pS6 


```{r, message = FALSE, warning = FALSE}
plot_s6 <- 
  function(cell_population, my_stimulation, my_condition = "dx") {
    
    patient_order <- 
      aml_data %>% 
      filter(
        condition == my_condition, 
        mahalanobis_cluster == cell_population, 
        stimulation %in% c(my_stimulation, "Basal")
      ) %>% 
      group_by(outcome, patient) %>% 
      filter(n() > 50) %>% 
      summarize(pS6 = median(pS6)) %>% 
      arrange(desc(outcome), desc(pS6)) %>% 
      pull(patient)
    
    aml_data %>% 
      filter(
        condition == my_condition,
        mahalanobis_cluster == cell_population, 
        stimulation %in% c(my_stimulation, "Basal")
      ) %>% 
      group_by(patient) %>% 
      filter(n() > 50) %>% 
      mutate(patient = factor(patient, levels = patient_order)) %>% 
      ggplot(aes(x = pS6, y = patient, fill = stimulation, linetype = outcome)) + 
      geom_density_ridges(alpha = 0.5) + 
      labs(caption = str_c("pS6 in ", cell_population, "-like cells"))
  } 

```


The phenograph paper found that pS6 responses to various stimulations differed by cell type in healthy samples, so can confirm that in our data as well. 

```{r}
feature_names %>% 
  filter(marker == "pS6") %>% 
  mutate(plots = map2(.x = cell_population, .y = stimulation, .f = plot_s6, my_condition = "healthy")) %>% 
  pull(plots) %>% 
  walk(.x = ., .f = print)

```

```{r}

my_stimulations <- 
  feature_names %>% 
  filter(marker == "pS6") %>% 
  pull(stimulation) %>% 
  unique()

patient_order <- 
      aml_data %>% 
      filter(
        condition == "healthy", 
        stimulation %in% c(my_stimulations, "Basal")
      ) %>% 
      group_by(patient) %>% 
      filter(n() > 50) %>% 
      summarize(pS6 = median(pS6)) %>% 
      arrange(desc(pS6)) %>% 
      pull(patient)
    
    aml_data %>% 
      filter(
        condition == "healthy", 
        stimulation %in% c(my_stimulations, "Basal")
      ) %>% 
      group_by(patient) %>% 
      filter(n() > 50) %>% 
      mutate(mahalanobis_cluster = factor(mahalanobis_cluster, levels = rev(CLASSIFIER_POPULATIONS))) %>% 
      ggplot(aes(x = pS6, y = mahalanobis_cluster, fill = stimulation)) + 
      geom_density_ridges(alpha = 0.5)

feature_names %>% 
  filter(marker == "pS6") %>% 
  mutate(plots = map2(.x = cell_population, .y = stimulation, .f = plot_s6)) %>% 
  pull(plots) %>% 
  walk(.x = ., .f = print)
  
```










