---
title: "Master AML Vignette"
author: 
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries-and-params, message=FALSE, warning=FALSE}
# Libraries
library(flowCore)
library(tidyverse)
library(readxl)
library(ggridges)
library(rlang)
library(ggthemes)
library(DataExplorer)
library(ComplexHeatmap)

# Parameters
source(here::here("scripts", "setup", "aml_utils.R")) #may need to change on scg machines
set_global_variables(locale = "galaxia")

tidyTOF_directory <- file.path("~", "GitHub", "tidyTOF")
is_sampled <- TRUE
md_path <- here::here("data-raw", "AML_metadata.xlsx")

# Sourcing tidyTOF functions
source_tidyTOF(tidyTOF_directory)

marker_setup()
patient_setup()

```

## Read in metadata

```{r read-metadata}
md <- 
  md_path %>% 
  read_excel() %>% 
  mutate(
    Patient = Patient %>% str_to_lower()
  ) %>% 
  janitor::clean_names(case = "snake")
```


## Read in single-cell data

### From scratch 

```{r raw-read-single-cell, read-raw-data, eval = FALSE}
source(file.path(CODE_DIRECTORY, "aml_read_data.R"))
```

### From .rds file

```{r rds-read-single-cell}
data_path <- 
  file.path(
    DATA_DIRECTORY, 
    if_else(is_sampled, "raw_aml_tibble_sampled.rds", "raw_aml_tibble.rds")
  )

aml_data <- 
  data_path %>% 
  read_rds() %>% 
  ungroup() %>% 
  mutate(cell_id = as.character(1:n()))
```


### Compare metadata and single-cell data

```{r compare-patient-names-pre}
print(
  str_glue(
    "Patients in MD but not in single-cell data: \n {my_patients} \n",
    my_patients = str_c(setdiff(md$patient,str_to_lower(aml_data$patient)), collapse = " ")
  )
)


print(
  str_glue(
    "Patients in single-cell data but not in MD: \n {my_patients} \n",
    my_patients = str_c(setdiff(str_to_lower(aml_data$patient), md$patient), collapse = " "), 
  )
)

```

Thus, we can see that several patients in the single-cell data set have to be renamed
in order to match the metadata. 

```{r fix-patient-names}
aml_data <- 
  aml_data %>% 
  mutate(
    patient = 
      recode(patient, PASBEE = "pasbbe", PASTRP = "pasrtp") %>% 
      str_to_lower(), 
    plate = factor(plate, levels = str_c("plate", 1:16))
  )
```


And now we check the names again... 
```{r compare-patient-names-post}
print(
  str_glue(
    "Patients in MD but not in single-cell data: \n {my_patients} \n",
    my_patients = str_c(setdiff(md$patient,str_to_lower(aml_data$patient)), collapse = " ")
  )
)


print(
  str_glue(
    "Patients in single-cell data but not in MD: \n {my_patients} \n",
    my_patients = str_c(setdiff(str_to_lower(aml_data$patient), md$patient), collapse = " "), 
  )
)
```

And we can see that there are some patients that included in the metadata that we have no single-cell data
for and that we have no metadata for the 4 healthy samples (the names that take the form `bm####`). 

### Pre-process

With arcsinh transformation and removal of added noise: 
```{r preprocess-1}
aml_data_p <- 
  aml_data %>% 
  tof_preprocess()
```

With arcsinh transformation and removal of added noise, plus scaling/centering: 

```{r preprocess-2}
ps_fun <- function(x) { 
  x %>% 
    {asinh(x = ./5)} %>% 
    scale() %>% 
    as.numeric()
}

aml_data_ps <- 
  aml_data %>% 
  tof_preprocess(transform_fun = ps_fun)
```

With removal of added noise and rank transformation (similar to "cytoDx"):
```{r preprocess-3}
aml_data_r <-
  aml_data %>%
  tof_preprocess(transform_fun = rank)
```


Save pre-processed data

```{r save-preprocessed-data}
ls() %>% 
  str_subset(pattern = "aml_data_.") %>% 
  walk(
    .f = ~ get(.x) %>% 
      write_rds(
        path = 
          file.path(
            DATA_DIRECTORY, 
            str_c(if_else(is_sampled, "sampled_", ""), .x, ".rds")
          )
      )
  )
```


## Data quality and univariate distributions

### Introduction to the data

First, we use the `{{DataExplorer}}` package to get to know the data a bit. 

```{r}
plot_str(aml_data)
```

```{r}
introduce(aml_data) %>% 
  pivot_longer(
    cols = everything(), 
    names_to = "dataset_feature", 
    values_to = "value"
  ) %>% 
  knitr::kable()
```

Number of cells across all discrete features: 

```{r}
discrete_bars <- 
  plot_bar(aml_data, ggtheme = theme_bw(), nrow = 2L, ncol = 2L, parallel = TRUE) %>% 
  `$`(page_1) + 
  labs(y = "Number of cells")
```


Histograms for all marker channels: 

```{r}
plot_channels <- 
  aml_data_p %>%
  summarize(across(one_of(ALL_MARKERS), mean)) %>% 
  pivot_longer(cols = everything(), names_to = "channel", values_to = "mean") %>% 
  arrange(desc(mean)) %>% 
  pull(channel)

aml_data_p %>% 
  select(one_of(plot_channels)) %>%
  plot_histogram(geom_histogram_args = list(bins = 40L))
```

Correlation matrix for all protein measurements: 

```{r corr-matrix-euclidean}
aml_data_p %>% 
  select(one_of(plot_channels), -CD3_CD19, -`caspase-3`) %>% 
  cor() %>% 
  Heatmap(
    col = 
      circlize::colorRamp2(
        breaks = seq(-1, 1, 0.5), 
        RColorBrewer::brewer.pal(n = 5, name = "RdBu")
      ),
    heatmap_legend_param = list(title = "Correlation"), 
    width = unit(15, "cm"), 
    height = unit(15, "cm")
  )
```

```{r corr-matrix-pearson}
aml_data_p %>% 
  select(one_of(plot_channels), -CD3_CD19, -`caspase-3`) %>% 
  cor() %>% 
  Heatmap(
    col = 
      circlize::colorRamp2(
        breaks = seq(-1, 1, 0.5), 
        RColorBrewer::brewer.pal(n = 5, name = "RdBu")
      ),    
    clustering_distance_rows = "pearson", 
    clustering_distance_columns = "pearson",
    heatmap_legend_param = list(title = "Correlation"), 
    width = unit(15, "cm"), 
    height = unit(15, "cm")
  )
```

```{r corr-matrix-manhattan}
aml_data_p %>% 
  select(one_of(plot_channels), -CD3_CD19, -`caspase-3`) %>% 
  cor() %>% 
  Heatmap(
    col = 
      circlize::colorRamp2(
        breaks = seq(-1, 1, 0.5), 
        RColorBrewer::brewer.pal(n = 5, name = "RdBu")
      ),
    clustering_distance_rows = "manhattan", 
    clustering_distance_columns = "manhattan",
    heatmap_legend_param = list(title = "Correlation"), 
    width = unit(15, "cm"), 
    height = unit(15, "cm")
  )
```

### By plate

#### Number of cells

```{r cell-number-plate}
aml_data %>% 
  count(plate) %>% 
  knitr::kable()
```

```{r cell-number-plate-density}
aml_data %>% 
  count(plate) %>% 
  ggplot(aes(x = n)) + 
  geom_density() + 
  geom_point(aes(x = n, y = 0), size = 3, shape = "|") + 
  scale_x_continuous(
    labels = scales::label_number(accuracy = 1, scale = 1e-6, suffix = "M")
  ) +
  scale_y_continuous() + 
  labs(
    subtitle = "Number of cells on each CyTOF plate", 
    caption = "Hash marks indicate individual plate values", 
    x = "Number of cells", 
    y = "Relative Density"
  )
```

#### Univariate histograms

```{r histograms-plate-plot, eval = FALSE}
aml_data_p %>% 
  tof_plot_all_histograms(
    group_var = plate, 
    out_path = file.path(HISTOGRAM_OUTPUT, "p"), 
    label_prefix = str_c(if_else(is_sampled, "sampled_", ""), "aml_data_p"),  
    label_suffix = "hist_by_plate.pdf", 
    width = 6,
    height = 10,
    device = "pdf"
  )

aml_data_p %>% 
  filter(condition == "healthy") %>% 
  tof_plot_all_histograms(
    group_var = plate, 
    out_path = file.path(HISTOGRAM_OUTPUT, "p"), 
    label_prefix = str_c(if_else(is_sampled, "sampled_", ""), "aml_data_p_healthy"), 
    label_suffix = "hist_by_plate.pdf", 
    width = 6, 
    height = 10,
    device = "pdf"
  )

aml_data_ps %>% 
  tof_plot_all_histograms(
    group_var = plate, 
    out_path = file.path(HISTOGRAM_OUTPUT, "ps"), 
    label_prefix = str_c(if_else(is_sampled, "sampled_", ""), "aml_data_ps"), 
    label_suffix = "hist_by_plate.pdf", 
    width = 6, 
    height = 10,
    device = "pdf"
  )

aml_data_ps %>% 
  filter(condition == "healthy") %>% 
  tof_plot_all_histograms(
    group_var = plate, 
    out_path = file.path(HISTOGRAM_OUTPUT, "ps"), 
    label_prefix = str_c(if_else(is_sampled, "sampled_", ""), "aml_data_ps_healthy"), 
    label_suffix = "hist_by_plate.pdf", 
    width = 6, 
    height = 10,
    device = "pdf"
  )

```

#### By plate (comparing patients on more than one plate)

```{r, warning = FALSE, message = FALSE}
my_patients <- 
  aml_data %>% 
  count(patient, plate) %>% 
  select(-n) %>% 
  count(patient) %>% 
  filter(n > 1) %>% 
  pull(patient)

aml_data_p %>% 
  filter(patient %in% my_patients) %>% 
  mutate(patient_plate = str_c(patient, plate, sep = "_")) %>% 
  tof_histogram(channel_var = CD45, group_var = patient_plate, ordered = TRUE) + 
  labs(y = NULL)

aml_data_p %>% 
  filter(patient %in% my_patients) %>% 
  mutate(patient_plate = str_c(patient, plate, sep = "_")) %>% 
  tof_histogram(channel_var = pSTAT5, group_var = patient_plate, ordered = TRUE) + 
  labs(y = NULL)

aml_data_p %>% 
  filter(patient %in% my_patients) %>% 
  mutate(patient_plate = str_c(patient, plate, sep = "_")) %>% 
  tof_histogram(channel_var = CD34, group_var = patient_plate, ordered = TRUE) + 
  labs(y = NULL)

aml_data_p %>% 
  filter(patient %in% my_patients) %>% 
  mutate(patient_plate = str_c(patient, plate, sep = "_")) %>% 
  tof_histogram(channel_var = CD38, group_var = patient_plate, ordered = TRUE) + 
  labs(y = NULL)
```


### By patient

#### Cell counts

```{r counts-patient}
aml_data %>% 
  count(patient) %>% 
  knitr::kable()
```


```{r cell-number-patient-density}
aml_data %>% 
  count(patient) %>% 
  ggplot(aes(x = n)) + 
  geom_density() + 
  geom_point(aes(x = n, y = 0), size = 3, shape = "|") + 
  scale_x_continuous(
    labels = scales::label_number(accuracy = 1, scale = 1e-3, suffix = "K")
  ) +
  scale_y_continuous() + 
  labs(
    subtitle = "Number of cells collected from each patient", 
    caption = "Hash marks indicate individual patient values", 
    x = "Number of cells", 
    y = "Relative Density"
  )
```

#### Univariate histograms
```{r histograms-patient, eval = FALSE}

aml_data_p %>% 
  tof_plot_all_histograms(
    group_var = patient, 
    out_path = file.path(HISTOGRAM_OUTPUT, "p"), 
    label_prefix = str_c(if_else(is_sampled, "sampled_", ""), "aml_data_p"), 
    label_suffix = "hist_by_patient.pdf", 
    width = 6, 
    height = 10,
    device = "pdf",
    ordered = TRUE
  )


aml_data_ps %>% 
  tof_plot_all_histograms(
    group_var = patient, 
    out_path = file.path(HISTOGRAM_OUTPUT, "ps"), 
    label_prefix = str_c(if_else(is_sampled, "sampled_", ""), "aml_data_ps"), 
    label_suffix = "hist_by_patient.pdf", 
    width = 6, 
    height = 10,
    device = "pdf", 
    ordered = TRUE
  )

```


### By patient (comparing patients with both Dx and Rx)

#### Cell counts

```{r}
aml_data %>% 
  filter(patient %in% PAIRED_PATIENTS) %>% 
  count(patient, condition)
```

#### Univariate histograms

```{r signaling-condition-density-plots, message = FALSE, warning = FALSE}
SIGNALING_MARKERS %>% 
  syms() %>% 
  map(
    ~ 
      tof_histogram_2(
        filter(aml_data_p, patient %in% PAIRED_PATIENTS), 
        channel_var = !!., 
        group_var = patient, 
        split_var = condition, 
        alpha = 0.5, 
        panel_scaling = FALSE
      ) + 
      facet_grid(cols = vars(stimulation))
  ) %>% 
  walk(print)
```

### By stimulation (surface)

```{r, warning = FALSE, message = FALSE}
surface_plots <- 
  SURFACE_MARKERS %>% 
  syms() %>% 
  map(
    ~ 
      tof_histogram_2(
        aml_data_p, 
        channel_var = !!., 
        group_var = patient, 
        split_var = stimulation, 
        alpha = 0.5,
        ordered = TRUE,
        panel_scaling = FALSE
      ) 
  )

surface_plots %>% 
  walk(print)

legend <- surface_plots[[1]] %>% get_legend()

surface_plots[1:9] %>% 
  map(~ . + theme(legend.position = "none")) %>% 
  plot_grid(plotlist = ., align = "vh", ncol = 3) %>% 
  plot_grid(., legend, nrow = 1, rel_widths = c(3, 0.3))

```


#### Cell counts

```{r cell-number-stims}
aml_data %>% 
  count(stimulation) %>% 
  knitr::kable()
```


#### Univariate histograms

```{r}
aml_data_p %>% 
  tof_histogram(channel_var = CD45, group_var = stimulation, ordered = TRUE)

aml_data_p %>% 
  tof_plot_all_histograms(
    group_var = stimulation, 
    out_path = file.path(HISTOGRAM_OUTPUT, "p"), 
    label_prefix = str_c(if_else(is_sampled, "sampled_", ""), "aml_data_p"), 
    label_suffix = "hist_by_stim.pdf", 
    width = 6, 
    height = 10,
    device = "pdf",
    ordered = TRUE
  )

aml_data_p %>% 
  filter(condition == "healthy") %>% 
  tof_plot_all_histograms(
    group_var = stimulation, 
    out_path = file.path(HISTOGRAM_OUTPUT, "p"), 
    label_prefix = str_c(if_else(is_sampled, "sampled_", ""), "aml_data_p_healthy"), 
    label_suffix = "hist_by_stim.pdf", 
    width = 6, 
    height = 10,
    device = "pdf"
  )

aml_data_ps %>% 
  tof_plot_all_histograms(
    group_var = stimulation, 
    out_path = file.path(HISTOGRAM_OUTPUT, "ps"), 
    label_prefix =  str_c(if_else(is_sampled, "sampled_", ""), "aml_data_ps"), 
    label_suffix = "hist_by_stim.pdf", 
    width = 6, 
    height = 10,
    device = "pdf", 
    ordered = TRUE
  )

aml_data_ps %>% 
  filter(condition == "healthy") %>% 
  tof_plot_all_histograms(
    group_var = stimulation, 
    out_path = file.path(HISTOGRAM_OUTPUT, "ps"), 
    label_prefix =  str_c(if_else(is_sampled, "sampled_", ""), "aml_data_ps_healthy"), 
    label_suffix = "hist_by_stim.pdf", 
    width = 6, 
    height = 10,
    device = "pdf"
  )
```



## Classifier 

```{r}

```

## Dimensionality reduction

```{r}
pca_recipe <- 
  recipe(~., data = aml_data_p) %>% 
  step_zv(all_numeric()) %>% 
  step_normalize(all_numeric()) %>% 
  step_pca(all_numeric(), num_comp = 5) 

pca_prep <- prep(pca_recipe)

pca_result <- juice(pca_prep)


```



