---
title: "AML Feature Extraction Vignette"
author: tkeyes
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Libraries
libraries <- 
  c(
    "flowCore", 
    "tidyverse", 
    "readxl", 
    "ggridges", 
    "rlang", 
    "ggthemes", 
    "foreach", 
    "doParallel", 
    "ggrepel", 
    "ggiraph", 
    "ggiraphExtra",
    "emdist",
    "philentropy", 
    "FlowSOM", 
    "tidytof"
  )

source(here::here("scripts", "setup", "aml_utils.R"))
call_libraries(libraries)
source(here::here("scripts", "feature_extraction", "feature_extraction_utils.R"))


# Parameters
set_global_variables(locale = "galaxia")
md_path <- here::here("data-raw", "AML_metadata.xlsx")

CLUSTER_OUTPUT <- here::here("data", "cancer_data_clustered.rds")

# set up aml marker and patient information
marker_setup()
patient_setup()

# Misc globals for running different parts of this vignette
is_sampled <- FALSE
aml_path <- here::here("data", "aml_data_clustered.rds")
all_path <- here::here("data", "all_data_clustered.rds")
sample_string <- ""

if (is_sampled) { 
  aml_path <- aml_path %>% 
    str_replace(".rds", "_sampled.rds")
  all_path <- all_path %>% 
    str_replace(".rds", "_sampled.rds")
  sample_string <- "_sampled"
}

```

# Executive Summary 

In this document, we formalize the feature extraction methods used in a variety of CyTOF papers as a preamble to using them in some predictive models. The feature extraction methods that we use include the following: 

* [Developmentally-Dependent Predictor of Relapse (DDPR)](https://pubmed.ncbi.nlm.nih.gov/29505032/)
* [CITRUS](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4084463/)/[Statistical Scaffold](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4537647/)
* Some hybrid methods using [Earth-Mover's Distance](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4805242/) or  the [Jensen-Shannon divergence](https://en.wikipedia.org/wiki/Jensen%E2%80%93Shannon_divergence) for signaling features (which was previously used in [this paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4076922/))

# Read in data

```{r}
aml_data <-
  aml_path %>%
  read_rds()

all_data <-
  all_path %>%
  read_rds()
```

If we want to use sampled data (for development), we take 10% of the cells that were read in from each file.

```{r eval=FALSE, include=FALSE}
if (is_sampled) { 
  aml_data <-
    aml_data %>%
    group_by(file_names) %>%
    slice_sample(prop = 0.05) %>%
    ungroup()
  
  all_data <-
    all_data %>%
    group_by(file_name) %>%
    slice_sample(prop = 0.05) %>%
    ungroup()
  }
```


# DDPR 

The DDPR feature matrix has 3 types of features:

* Proportion of cells in each cell population *expanded* relative to healthy controls.
* Mean expression of lineage markers in each of these sub-populations
* Proportion of cells that are "positive" for each of the measured signaling markers (which we implement as being above a certain threshold). 

To get to each of these feature types, we can create some functions that get us most of the way there and then wrap them in an overall `tof_extract_features` function that allows us to specify ddpr as the method we want to use. 

## Helper functions

First, we source some helper functions that haven't yet been incorporated into `tidytof.`

```{r}
source(here::here("scripts", "feature_extraction", "feature_extraction_utils.R"))
```


## Calculate DDPR feature matrices

The first step to compute the DDPR feature matrices is to separate the cells that come from healthy samples and those that come from cancer samples. 

```{r}
healthy_tibble <- 
  aml_data %>% 
  filter(condition == "healthy")


cancer_tibble <- 
  aml_data %>% 
  filter(condition != "healthy") %>% 
  group_by(patient, condition)

```


Then, we can find the DDPR feature matrices for each patient and condition using a variety of signaling thresholds using the helper function `aml_iterate_ddpr_extraction` in `aml_utils.R`. 

We do this below using a variety of clustering techniques to define the clusters from which we are extracting ddpr features including the following: 

- developmental classification (using mahalanobis distance)
- flowsom clustering (using surface markers)
- flowsom clustering using the mahalanobis distances from single-cell classification
- flowsom clustering using the cosine distances from single-cell classification

The result, a tibble with a list-col containing the extracted features, will be saved to our data folder. 

```{r, warning = FALSE, message = FALSE}
ddpr_feature_matrices <- 
  tibble(
    threshold = asinh(c(5, 10, 20, 50, 100)), 
    features_mahalanobis = 
      map(
        .x = threshold, 
        .f = ~ 
          aml_iterate_ddpr_extraction(
            cluster_col = mahalanobis_cluster, 
            threshold = .x
          )
      ),
    features_cos =
      map(
        .x = threshold, 
        .f = ~ 
          aml_iterate_ddpr_extraction(
            cluster_col = cosine_cluster, 
            threshold = .x
          )
      ),
    features_flowsom_surface = 
      map(
        .x = threshold, 
        .f = ~ 
          aml_iterate_ddpr_extraction(
            cluster_col = cluster_flowSOM_surface, 
            threshold = .x
          )
      ),
    features_flowsom_mah = 
      map(
        .x = threshold, 
        .f = ~ 
          aml_iterate_ddpr_extraction(
            cluster_col = cluster_flowSOM_mah, 
            threshold = .x
          )
      ), 
    features_flowsom_cos = 
      map(
        .x = threshold, 
        .f = ~ 
          aml_iterate_ddpr_extraction(
            cluster_col = cluster_flowSOM_cos, 
            threshold = .x
          )
      )
  ) %>% 
  pivot_longer(
    cols = starts_with("features_"), 
    names_to = "cluster_type", 
    values_to = "features", 
    names_prefix = "features_"
  )



ddpr_feature_matrices

file_name <- str_c("ddpr_feature_matrices", sample_string, ".rds")

write_rds(
  x = ddpr_feature_matrices, 
  file = 
    here::here(
      "data", 
      "patient-level_feature_matrices", 
      file_name
    )
)

```

## Calculate abundance + central tendency feature matrices

We can also extract feature matrices from our single-cell data that simply extract the mean of each of our markers in each clustered cell type within each stimulation condition. For good measure, we can also include the proportion of each cluster in these matrices. 

For this, we will use the same 4 clustering strategies as above. First we start with finding the means...

```{r}
mean_feature_matrices <-
  tibble(
    features_ddpr_mahalanobis = 
      list(
        aml_extract_lineage_features(
          cluster_col = mahalanobis_cluster, 
          central_tendency = mean
        )
      ),
    features_ddpr_cos = 
      list(
        aml_extract_lineage_features(
          cluster_col = cosine_cluster, 
          central_tendency = mean
        )
      ), 
    features_flowsom_surface = 
      list(
        aml_extract_lineage_features(
          cluster_col = cluster_flowSOM_surface, 
          central_tendency = mean
        )
      ),
    features_flowsom_mah = 
      list(
        aml_extract_lineage_features(
          cluster_col = cluster_flowSOM_mah, 
          central_tendency = mean
        )
      ), 
    features_flowsom_cos = 
      list(
        aml_extract_lineage_features(
          cluster_col = cluster_flowSOM_cos, 
          central_tendency = mean
        )
      )
  ) %>% 
  pivot_longer(
    cols = starts_with("features_"), 
    names_to = "cluster_type", 
    values_to = "features", 
    names_prefix = "features_"
  )

    
mean_feature_matrices
```


...and then we find the abundances.

```{r}
abundance_feature_matrices <-
  tibble(
    features_ddpr_mahalanobis = 
      list(
        extract_feature_proportion(
          tof_tibble = group_by(aml_data, patient, condition),
          cluster_col = mahalanobis_cluster
        )
      ),
    features_ddpr_cos = 
      list(
        extract_feature_proportion(
          tof_tibble = group_by(aml_data, patient, condition), 
          cluster_col = cosine_cluster
        )
      ),
    features_flowsom_surface = 
      list(
        extract_feature_proportion(
          tof_tibble = group_by(aml_data, patient, condition),
          cluster_col = cluster_flowSOM_surface 
        )
      ),
    features_flowsom_mah = 
      list(
        extract_feature_proportion(
          tof_tibble = group_by(aml_data, patient, condition),
          cluster_col = cluster_flowSOM_mah 
        )
      ), 
    features_flowsom_cos = 
      list(
        extract_feature_proportion(
          tof_tibble = group_by(aml_data, patient, condition),
          cluster_col = cluster_flowSOM_cos 
        )
      )
  ) %>% 
  pivot_longer(
    cols = starts_with("features_"), 
    names_to = "cluster_type", 
    values_to = "features", 
    names_prefix = "features_"
  ) %>%
  rename(abundances = features)


```

And then we combine the mean and abundance features:

```{r}
final_features <- 
  left_join(
    mean_feature_matrices, 
    abundance_feature_matrices, 
    by = "cluster_type"
  ) %>% 
  transmute(
    cluster_type, 
    final_features = 
      map2(features, abundances, left_join, by = c("patient", "condition"))
  )


file_name <- str_c("mean_feature_matrices", sample_string, ".rds")

write_rds(
  x = final_features, 
  file = 
    here::here(
      "data", 
      "patient-level_feature_matrices", 
      file_name
    )
)
```

## Calculate "CITRUS" feature matrices

The citrus feature matrix is similar to the one taken directly above, but takes into account which stimulation condition each cell is in (and separates each stimulation condition into its own unique feature for each cell population and marker). 

For now, we won't worry about calculating the CITRUS feature matrix because with such a small number of samples, I am worried about overfitting (with a very high number of features)

```{r}
citrus_feature_matrices <- NULL
```



## Calculate feature matrices (with EMD for signaling)

Instead of calculating a measurement of central tendency for each signaling feature (for each stimulation), can also look at measures of how different the entire distributions are between the basal condition and each of the stimulation conditions. 

One way of doing this is to calculate the Earth Mover's Distance(EMD) between the distribution of cells in the basal condition and the distribution of cells in each stimulation condition within a given patient. This will yield a "score" measuring how different the distributions (one score for each non-basal stimulation condition).

Note that calculating the EMD is very computationally expensive. A single EMD calculation for one set of clusters (on heavily sampled data) takes about 14 minutes locally.


```{r, eval = FALSE}
emd_feature_matrices <- 
  list(
    ddpr_mahalanobis = 
      aml_extract_lineage_emd(mahalanobis_cluster, mean) %>% 
      ungroup(), 
    ddpr_cos = 
      aml_extract_lineage_emd(cosine_cluster, mean) %>% 
      ungroup(), 
    flowsom_surface = 
      aml_extract_lineage_emd(cluster_flowSOM_surface, mean) %>% 
      ungroup(), 
    flowsom_mah = 
      aml_extract_lineage_emd(cluster_flowSOM_mah, mean) %>% 
      ungroup(), 
    flowsom_cos = 
      aml_extract_lineage_emd(cluster_flowSOM_cos, mean) %>% 
      ungroup()
  )

final_features <- 
  tibble(
    cluster_type = names(emd_feature_matrices), 
    features = emd_feature_matrices
  )

file_name <- str_c("emd_feature_matrices", sample_string, ".rds")


write_rds(
  x = final_features, 
  file = 
    here::here(
      "data", 
      "patient-level_feature_matrices", 
      file_name
    )
)

```



### Calculate feature matrices with JS Index

Another measure of the distance between two distributions is the Jensen-Shannon index. We can calculate it for all of our samples similar to how we calculated EMD...

The JSI is not quite as computationally expensive as the EMD.

```{r}
js_feature_matrices <- 
  list(
    ddpr_mahalanobis = 
      aml_extract_lineage_js(mahalanobis_cluster, mean) %>%
      ungroup(),
    ddpr_cos =
      aml_extract_lineage_js(cosine_cluster, mean) %>%
      ungroup(),
    flowsom_surface = 
      aml_extract_lineage_js(cluster_flowSOM_surface, mean) %>% 
      ungroup(),
    flowsom_mah =
      aml_extract_lineage_js(cluster_flowSOM_mah, mean) %>%
      ungroup(),
    flowsom_cos =
      aml_extract_lineage_js(cluster_flowSOM_cos, mean) %>%
      ungroup()
  )

js_tibble <- 
  tibble(
    cluster_type = names(js_feature_matrices), 
    features = js_feature_matrices
  )

js_tibble


file_name <- str_c("js_feature_matrices", sample_string, ".rds")


write_rds(
  x = js_tibble,
  file =
    here::here(
      "data",
      "patient-level_feature_matrices",
      file_name
    )
)
```



