---
title: AML project and data analysis pipeline report
author: 
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE, include = FALSE}
# Libraries
library(tidyverse)

# Parameters

#===============================================================================

# Code
```

# A systematic approach to CyTOF modeling

## Reading data and metadata 

A standardized analysis pipeline has to be able to read in data from a standard format
and into a readily-useable data structure.

Associated tidyTOF functions:

* `tof_read_fcs()` - to read in .fcs files in bulk (1.5x speed improvement over `{cytofkit}`)
* `tof_clean_marker_names()` - format names initially assigned to each channel by the mass cytometer
* `tof_read_metadata()` - read in a .csv or .xls file re: patient-level metadata for each sample

## Homogenization of datasets

Associated tidyTOF functions:

* `ben_channel_homogenization()` - still under development

## Data Pre-processing

Tasks:

* Variance-stabilizing transformation (`tof_arcsinh()`)
* "Normalization" of raw data
  * Percentile normalization
  * Quantile normalization
  * Rank-order transformation
  * MMD sample-to-sample normalization (SAUCIE and DeepCyTOF)
  * Batch normalization (Ben)
  * Others?

## Feature engineering 

* Patient-level matrix extraction
  * `tof_extract_feature_matrix()`
    * Statistical Scaffold/CITRUS
    * DDPR 
    * CytoDx
  * Developmental embedding
    * Neural-network-based (`word2vec`)
    * Path-tracing algorithm (monocle, slingshot)
* Single-cell level
  * “Eigencells” across stimulation conditions using knn on surface markers 

## Modeling

* GLMs (`{tidymodels}`) 
* Neural networks (`{keras}` and `{tensorflow}`)
